Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2021-11-13 docs/00-manuscripts/mdpi/manuscript.tex
clarify confusing sentence

Some individual features were removed before using the datasets for modeling when being numerically equivalent to another feature based on the pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates for feature set~VI to 86 (from 89).
This decision was made to prevent issues for the subsequent tuning, filtering and model fitting steps which might occur when passing features with a pairwise correlation of (almost) one.
Additional variable selection was solely devoted to the respective feature selection methods in this study. <!-- FIXME: I don't understand this sentence. -->

@@

% feature sets
% - no substantial difference between feature sets
% - combining feature sets has only small impact
% - expert based feature sets show no better result than data-driven ones
% - penalized methods seem to have problems with VI feature set
The combination of feature sets showed small increases in performance for some learners.
RF and XGBoost scored slightly better on the combined datasets HR-NRI, NRI-VI and HR-NRI-VI, respectively, compared to their standalone variants (NRI, VI) (\autoref{fig:perf-result}).
However, the best performances for both RF and XGBoost were scored on HR only.
Datasets containing derived features only (VI, NRI) showed no improvement in performance compared to the raw hyperspectral band information (HR) or combined feature sets.

% learner
% - SVM Car is the overall winner
% - Lasso quite stable (varies within 1 - 4), ridge a bit more variable (2 - 6)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Relief} filter achieved the best overall performance (RMSE of 28.09 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalties showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Differences to other feature sets were very small for lasso (below one p.p.) and a bit larger for ridge regression (between two and five p.p.).
XGBoost showed very poor performances for some feature sets and fills the last\ ten places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
The use of filters with RF resulted in a substantial increase in performance in all tasks with the exception of NRI, in which the difference between all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost's performance depended strongly on feature selection: In two out of six tasks, using no filter resulted in the worst or second\-/worst performance.
XGBoost showed the highest overall differences between filters for a single task: for feature set HR, the range is up to 15 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there were no instances where a model without filtering scored a better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and \enquote{no filter} achieved roughly the same performance on all tasks.

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter achieved in only one instance (RF on HR-NRI-VI) the best performance of\ all filters (\autoref{fig:filter-effects-borda}).
For RF and XGBoost, it most often ranked within the better half among all filters of a specific task.
For XGBoost on the NRI task, the Borda filter scored the second\-/worst performance.

% Selected percentage of features across plots
The number of features selected based on model tuning strongly varied across learners and plots.
While for RF, the least features were selected when Luiando or Oiartzun was the test set ($n = 1$), more than 90\% of all features were used for test sets~Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots and among all learners, the highest\ single fraction being 24\% for test set~Laukiz2.
In general, all learners behaved rather differently on the various plots, and therefore no overall pattern could be identified.

@@

The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For the HR dataset, most (i.e., six out of ten) relevant features clustered around the infrared region (920~nm - 1000~nm), while for VI, eight out of ten concentrate within the wavelength range of 700~nm - 750~nm (the so\-/called \enquote{red edge}).
For HR, four features in the infrared region (920~nm - 1000~nm) were identified by the model to be most important, being associated with a mean decrease in RMSE of around 1 p.p.
Overall, apart from the top five features, the vast majority of features showed only a small importance with average decreases in RMSE below 0.5 p.p.

@@

The best aggregated performance of this study (SVM + \enquote{Relief} filter, RMSE~28.12 p.p.) has to be seen in the light of model overfitting (see \autoref{subsec:perf-plot-char}).
Leaving out the performance on Laukiz2 when aggregating results, the mean RMSE would be around 19 p.p.
However, leaving out a single plot would also change the prediction results for the other plots because the observations from Laukiz2 would not be available for model training.
Due to the apparent presence of model overfitting in this study, it can be postulated that more training data representing a greater variety of situations is needed.
A model can only make robust predictions if it has learned relationships across the whole range of the response.
Hence, care should be taken when predicting to the landscape scale using models fitted on this dataset due to their lack of generalizability caused by the limitations of the available training data.
However, when inspecting the fold\-/level performances, it can be concluded that the models performed reasonably well predicting defoliation greater than 50\% but failed for lower levels.
This applied to all learners of this study.
The overall performance of all learners achieved in this study can be classified as ``poor'' given that only the SVM learner was able to outperform the featureless learner (\autoref{tab:best-learner-perf}).

@@

Arbitrary seeming values of selected feature percentages like 99.97268\% in \autoref{tab:tune-perc-sel-features} emerge from the specific setting which got elected as the best setting during the tuning step.
The specific value to benchmark during tuning is defined by the internal surrogate learner of the MBO tuning method in this study and can be any numeric value between 0 and 1 (interpreted as a\ percentage of available features).

@@

% - Ensemble FS shows no advantage to simple methods -> always one simple filter which is better. Ensemble filters are good across many different datasets
The models which used the Borda ensemble method in this study did not score better on average than models which used a single filter or no filter at all.
Ensemble methods have higher stability and robustness than single ones and have shown promising results in \cite{drotar2017}.
Hence, their main advantage is stable performances across datasets with varying characteristics.
Single filter methods might yield better model performances on certain datasets but fail on others.
The fact that this study used multiple feature sets but only one dataset and tested many single filters could be a potential explanation why in all cases (besides RF on task HR-NRI-VI) a single filter outperformed the ensemble filter.
However, studies which used ensemble filters are still rare and usually these are not compared against single filters \cite{ghosh2019}.
In summary, Borda does\ not\ perform better than a randomly selected filter method in this study.
More case studies applying ensemble filter methods are needed to verify this finding.
Nevertheless, ensemble filters can be a promising addition to a ML feature-selection portfolio.

% - PCA shows similar performance but probably requires tuning of main components -> check runtime!
PCA, acting as a filter method, more often showed less optimal results, especially for algorithms RF and XGBoost.
Especially XGBoost had substantial problems when using PCA as a filter method and scored the four\ worst results (\autoref{tab:perf-worst-10}).
However, PCA was able to reduce model fitting times substantially across all algorithms.
Depending on the use case, PCA can be an interesting option to reduce dimensionality while keeping runtime low.
However, information about the total number of features used by the model is lost when applying this technique.
Since filter scores only need to be calculated once for a given dataset in a benchmark setting, the runtime advantage of a PCA vs. filter methods might in fact be negligible in practice.

>>>>

Some individual features were removed before using the datasets for modeling when being numerically equivalent to another feature based on the pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates for feature set~VI to 86 (from 89).
This decision was made to prevent issues for the subsequent tuning, filtering and model fitting steps which might occur when passing features with a pairwise correlation of (almost) one.
The remaining features were then used as input for the filter-based feature selection within the CV.

@@

% feature sets
% - no substantial difference between feature sets
% - combining feature sets has only small impact
% - expert based feature sets show no better result than data-driven ones
% - penalized methods seem to have problems with VI feature set
The combination of feature sets showed small increases in performance for some learners.
XGBoost scored slightly better on the combined datasets HR-NRI, NRI-VI and HR-NRI-VI, respectively, compared to their standalone variants (NRI, VI) (\autoref{fig:perf-result}).
However, the best performances for RF and XGBoost were scored on NRI and HR, respectively.
RF showed a substantial performance increase when using only NRI compared to all other feature sets whereas for XGBoost the worst performances were associated with the VI and NRI only feature sets (\autoref{fig:perf-result}).

% learner
% - SVM Info Gain is the overall winner
% - Combined sets do not help for XBoost and Ridge
% - RF NRI quite strong
SVM combined with the \enquote{Information Gain} filter achieved the best overall performance (RMSE of 27.915 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalties showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Combining feature sets for lasso and ridge did not help to increase performance and while there was no subatantial difference for lasso, ridge scored around two percentage points on such.
XGBoost showed very poor performances for some feature sets and fills the last\ ten places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
The use of filters for RF resulted in a substantial increase in performance in all tasks, especially on the HR feature set where all filters showed an improved score compared to using no filter (\autoref{fig:filter-effects-no-filter}).
XGBoost's performance depended strongly on feature selection: In two out of six tasks (HR, VI), using no filter resulted in the worst performance.
XGBoost showed the highest overall differences between filters for a single task: for feature set HR, the range is up to 13 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there were no instances in which a non-filtered model scored a better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and \enquote{no filter} achieved roughly the same performance on all tasks.

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter was not able to score the best performance in any learner/filter setting (\autoref{fig:filter-effects-borda}).
For RF and XGBoost, it most often ranked within the better half among all filters of a specific task.

% Selected percentage of features across plots
FIXME:
The number of features selected based on model tuning strongly varied across learners and plots.
While for RF, the least features were selected when Luiando or Oiartzun was the test set ($n = 1$), more than 90\% of all features were used for test sets~Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots and among all learners, the highest\ single fraction being 24\% for test set~Laukiz2.
In general, all learners behaved rather differently on the various plots, and therefore no overall pattern could be identified.

@@

FIXME:
The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For the HR dataset, most (i.e., six out of ten) relevant features clustered around the infrared region (920~nm - 1000~nm), while for VI, eight out of ten concentrate within the wavelength range of 700~nm - 750~nm (the so\-/called \enquote{red edge}).
For HR, four features in the infrared region (920~nm - 1000~nm) were identified by the model to be most important, being associated with a mean decrease in RMSE of around 1 p.p.
Overall, apart from the top five features, the vast majority of features showed only a small importance with average decreases in RMSE below 0.5 p.p.

@@

The best aggregated performance of this study (SVM + \enquote{Info Gain} filter, RMSE~27.915 p.p.) has to be seen in the light of model overfitting (see \autoref{subsec:perf-plot-char}).
Leaving out the performance on Laukiz2 when aggregating results, the mean RMSE would be around 19 p.p.
However, leaving out a single plot would also change the prediction results for the other plots because the observations from Laukiz2 would not be available for model training.
Due to the apparent presence of model overfitting in this study, it can be postulated that more training data representing a greater variety of situations is needed.
A model can only make robust predictions if it has learned relationships across the whole range of the response.
Hence, care should be taken when predicting to the landscape scale using models fitted on this dataset due to their lack of generalizability caused by the limitations of the available training data.
However, when inspecting the fold\-/level performances, it can be concluded that the models performed reasonably well predicting defoliation greater than 50\% but failed for lower levels.
This applied to all learners of this study.
The overall performance of all learners achieved in this study can be classified as ``poor'' given that only the SVM learner was able to outperform the featureless learner (\autoref{tab:best-learner-perf}).

@@

FIXME:
Arbitrary seeming values of selected feature percentages like 99.97268\% in \autoref{tab:tune-perc-sel-features} emerge from the specific setting which got elected as the best setting during the tuning step.
The specific value to benchmark during tuning is defined by the internal surrogate learner of the MBO tuning method in this study and can be any numeric value between 0 and 1 (interpreted as a\ percentage of available features).

@@

% - Ensemble FS shows no advantage to simple methods -> always one simple filter which is better. Ensemble filters are good across many different datasets
The models which used the Borda ensemble method in this study did not score better on average than models which used a single filter or no filter at all.
Ensemble methods have higher stability and robustness than single ones and have shown promising results in \cite{drotar2017}.
Hence, their main advantage is stable performances across datasets with varying characteristics.
Single filter methods might yield better model performances on certain datasets but fail on others.
The fact that this study used multiple feature sets but only one dataset and tested many single filters could be a potential explanation why in all cases a single filter outperformed the ensemble filter.
However, studies which used ensemble filters are still rare and usually these are not compared against single filters \cite{ghosh2019}.
In summary, Borda does\ not\ perform better than a randomly selected filter method in this study.
More case studies applying ensemble filter methods are needed to verify this finding.
Nevertheless, ensemble filters can be a promising addition to a ML feature-selection portfolio.

% - PCA shows similar performance but probably requires tuning of main components -> check runtime!
PCA, acting as a filter method, more often showed less optimal results, especially for algorithms RF and XGBoost.
Especially XGBoost had substantial problems when using PCA as a filter method and scored four of the six worst results (\autoref{tab:perf-worst-10}).
However, PCA was able to reduce model fitting times substantially across all algorithms.
Depending on the use case, PCA can be an interesting option to reduce dimensionality while keeping runtime low.
However, information about the total number of features used by the model is lost when applying this technique.
Since filter scores only need to be calculated once for a given dataset in a benchmark setting, the runtime advantage of a PCA vs. filter methods might in fact be negligible in practice.
